---
title: "Machine Learning"
output: html_notebook
---

Load required packages

```{r}
library(tidyverse)
library(glmnet)
```


Load cleaned and combined data

```{r}
data = read_csv('Data/combined_stats.csv')
data
```

Define columns that will be used for prediction

```{r}
predictors <- c('Age', 'G', 'MP', 'FG', 'FGA', 'FG%', '3P',
                '3PA', '3P%', '2P', '2PA', '2P%', 'eFG%', 'FT', 'FTA', 'FT%', 'ORB',
                'DRB', 'TRB', 'AST', 'STL', 'BLK', 'TOV', 'PF', 'PTS', 'Year', 'W', 'L', 'W/L%', 'GB', 'PS/G',
                'PA/G', 'SRS')
```

Split data into training and testing sets

```{r}
train <- data %>% filter(Year < 2024)
test <- data %>% filter(Year == 2024)
train
test
```

Make multiple linear regression model using data

```{r}
mlr_model <- lm(train$Share ~ ., data = train %>% select(all_of(predictors)))
summary(mlr_model)
```

Make predictions on the test data and evaluate performance

```{r}
mlr_predictions <- predict(mlr_model, newdata = test)
actual_values <- test$Share
test_results <- test
test_results$`MLR Prediction` <- mlr_predictions
test_results <- test_results %>% arrange(desc(Share))
test_results
mlr_mse <- mean((mlr_predictions - actual_values)^2)
mlr_mse
```

Make ridge regression model

```{r}
x_ridge <- train %>% select(all_of(predictors)) %>% as.matrix()
y_ridge <- train$Share
ridge_model <- glmnet(x_ridge, y_ridge, alpha = 0)
```

Find best lambda for ridge regression model

```{r}
cv_ridge <- cv.glmnet(x_ridge, y_ridge, alpha = 0)
best_lambda_ridge <- cv_ridge$lambda.min
best_lambda_ridge
```

Make predictions using ridge regression model and evaluate performance

```{r}
ridge_predictions <- predict(ridge_model, s = best_lambda_ridge, newx = test %>% select(all_of(predictors)) %>% as.matrix()) %>% as.vector()
actual_values <- test$Share
test_results <- test_results %>% arrange(Player)
test_results$`Ridge Prediction` <- ridge_predictions
test_results <- test_results %>% arrange(desc(Share))
test_results
ridge_mse <- mean((ridge_predictions - actual_values)^2)
ridge_mse
```

Make lasso model

```{r}
x_lasso <- train %>% select(all_of(predictors)) %>% as.matrix()
y_lasso <- train$Share
lasso_model <- glmnet(x_lasso, y_lasso, alpha = 1)
```

Find best lambda for lasso model

```{r}
cv_lasso <- cv.glmnet(x_lasso, y_lasso, alpha = 1)
best_lambda_lasso <- cv_lasso$lambda.min
best_lambda_lasso
```

Make predictions using lasso model and evaluate performance

```{r}
lasso_predictions <- predict(lasso_model, s = best_lambda_lasso, newx = test %>% select(all_of(predictors)) %>% as.matrix()) %>% as.vector()
actual_values <- test$Share
test_results <- test_results %>% arrange(Player)
test_results$`Lasso Prediction` <- lasso_predictions
test_results <- test_results %>% arrange(desc(Share))
test_results
lasso_mse <- mean((lasso_predictions - actual_values)^2)
lasso_mse
```

Add voting ranks

```{r}
test_results <- test_results %>% arrange(desc(Share))
test_results <- test_results %>% mutate(Rk = seq_len(n()))
test_results <- test_results %>% arrange(desc(`MLR Prediction`))
test_results <- test_results %>% mutate(`MLR Rk` = seq_len(n()))
test_results <- test_results %>% arrange(desc(`Ridge Prediction`))
test_results <- test_results %>% mutate(`Ridge Rk` = seq_len(n()))
test_results <- test_results %>% arrange(desc(`Lasso Prediction`))
test_results <- test_results %>% mutate(`Lasso Rk` = seq_len(n()))
test_results <- test_results %>% arrange(desc(Share))
test_results
```
